title: "Final Project - Hydro Power Generation and Streamflow Analysis"
author: "Kaitlyn Elliott and Jon Nichols"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
## Background and Methods
In this project we set out to explore the relationship between hydro power generation and stream generation. To do this we looked at power generation data from the EPI and stream gage data from the USGS. For generation data we selected generation sites/dams so that half were from the western United States and half from the East. Stream gage data was selected upstream from dams and reservoirs to get an accurate representation of stream flow over time. Both power generation data and stream flow was analyzed for seasonality and monotonic trends using a seasonal man kendall test. This was to help understand if generation was increasing or decreasing as a result of stream flow. Although outside factors could also contribute increases or decreases in power generation.

## Set Up  

```{r}

getwd()

library(tidyverse)
library(lubridate)
library(dataRetrieval)
library(trend)
library(dataRetrieval)
library(cowplot)
library(readr)
library(zoo)
library(Kendall)
library(tseries)

theme_set(theme_classic())

```

## Data Processing 

```{r}
#Reading in stream data

#East Stream Gage Sites

gage_02143000<-readNWISdata(siteNumbers = "02143000", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_02071000<-readNWISdata(siteNumbers = "02071000", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_01563200<-readNWISdata(siteNumbers = "01563200", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_01054500<-readNWISdata(siteNumbers = "01054500", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_02163001<-readNWISdata(siteNumbers = "02163001", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

#West Stream Gage Sites

gage_06279940<-readNWISdata(siteNumbers = "06279940", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_13046000<-readNWISdata(siteNumbers = "13046000", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_09146200<-readNWISdata(siteNumbers = "09146200", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_09328990<-readNWISdata(siteNumbers = "09328990", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31") # this one is broken

gage_09408135<-readNWISdata(siteNumbers = "09408135", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

#Summarizing Data by Month

gage_01054500 <- gage_01054500 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_01563200<-gage_01563200 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_02071000<-gage_02071000 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_02143000<-gage_02143000 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_02163001<-gage_02163001 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_06279940<-gage_06279940 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_09146200<-gage_09146200 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

#gage_09328990<-gage_09328990 %>%
  #mutate(Discharge = X_00060_00003) %>%
  #select(Discharge, dateTime, site_no) %>%
  #mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  #group_by(Year, Month) %>%
  #summarize(average_monthly_discharge = mean(Discharge))

gage_09408135<-gage_09408135 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_13046000<-gage_13046000 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))
#Reading in generation data
High_Shoals_Generation <- read.csv("./Final Project Hydro Generation Data/High_Shoals_Hydro_(NC)_net_generation (1).csv")

Schoolfield_Generation <- read.csv("./Final Project Hydro Generation Data/Schoolfield_Dam_net_generation.csv")

Matson_Generation <- read.csv("./Final Project Hydro Generation Data/William_F_Matson_Generating_Station_net_generation.csv")

Otis_Generation <- read.csv("./Final Project Hydro Generation Data/Otis_Hydro_net_generation.csv")

Ware_Shoals_Generation <- read.csv("./Final Project Hydro Generation Data/Ware_Shoals_Hydro_Project_net_generation.csv")

Garland_Generation <- read.csv("./Final Project Hydro Generation Data/Garland_Canal_Power_Plant_net_generation.csv")

Chester_Generation <- read.csv("./Final Project Hydro Generation Data/Chester_Diversion_Hydroelectric_Project_net_generation.csv")

Tri_County_Generation <- read.csv("./Final Project Hydro Generation Data/Tri-County_Water_Hydropower_Project_net_generation.csv")

Glen_Canyon_Generation <- read.csv("./Final Project Hydro Generation Data/Glen_Canyon_Dam_net_generation.csv")

Quail_Creek_Generation <- read.csv("./Final Project Hydro Generation Data/Quail_Creek_Hydro_Plant_#1_net_generation.csv")


High_Shoals_Generation <- High_Shoals_Generation[-c(1),]
Schoolfield_Generation <- Schoolfield_Generation[-c(1),]
Matson_Generation <- Matson_Generation[-c(1),]
Otis_Generation <- Otis_Generation[-c(1),]
Ware_Shoals_Generation <-Ware_Shoals_Generation[-c(1),] 
Garland_Generation <- Garland_Generation[-c(1),]
Chester_Generation <- Chester_Generation[-c(1),]
Tri_County_Generation <- Tri_County_Generation[-c(1),]
Glen_Canyon_Generation <- Glen_Canyon_Generation[-c(1),]
Quail_Creek_Generation <- Quail_Creek_Generation[-c(1),]

colnames(High_Shoals_Generation) <- c("date_wrong", "Power")
colnames(Schoolfield_Generation) <- c("date_wrong", "Power")
colnames(Matson_Generation) <- c("date_wrong", "Power")
colnames(Otis_Generation) <- c("date_wrong", "Power")
colnames(Ware_Shoals_Generation) <- c("date_wrong", "Power")
colnames(Garland_Generation) <- c("date_wrong", "Power")
colnames(Chester_Generation) <- c("date_wrong", "Power")
colnames(Tri_County_Generation) <- c("date_wrong", "Power")
colnames(Glen_Canyon_Generation) <- c("date_wrong", "Power")
colnames(Quail_Creek_Generation) <- c("date_wrong", "Power")

High_Shoals_Generation <- High_Shoals_Generation %>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Schoolfield_Generation <- Schoolfield_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Matson_Generation <- Matson_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Otis_Generation <- Otis_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Ware_Shoals_Generation <-Ware_Shoals_Generation %>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Garland_Generation <- Garland_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Chester_Generation <- Chester_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Tri_County_Generation <- Tri_County_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Glen_Canyon_Generation <- Glen_Canyon_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Quail_Creek_Generation <- Quail_Creek_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))

High_Shoals_Generation <- High_Shoals_Generation %>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)
  

Schoolfield_Generation <- Schoolfield_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Matson_Generation <- Matson_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Otis_Generation <- Otis_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Ware_Shoals_Generation <-Ware_Shoals_Generation %>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Garland_Generation <- Garland_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Chester_Generation <- Chester_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Tri_County_Generation <- Tri_County_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Glen_Canyon_Generation <- Glen_Canyon_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Quail_Creek_Generation <- Quail_Creek_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)



High_Shoals_Generation_Complete <-  High_Shoals_Generation[!(is.na(High_Shoals_Generation$Power) | High_Shoals_Generation$Power==""), ]

Schoolfield_Generation_Complete <-  Schoolfield_Generation[!(is.na(Schoolfield_Generation$Power) | Schoolfield_Generation$Power==""), ]

Matson_Generation_Complete <-  Matson_Generation[!(is.na(Matson_Generation$Power) | Matson_Generation$Power==""), ]

Otis_Generation_Complete <-  Otis_Generation[!(is.na(Otis_Generation$Power) | Otis_Generation$Power==""), ]

Ware_Shoals_Generation_Complete <-  Ware_Shoals_Generation[!(is.na(Ware_Shoals_Generation$Power) | Ware_Shoals_Generation$Power==""), ]

Garland_Generation_Complete <-  Garland_Generation[!(is.na(Garland_Generation$Power) | Garland_Generation$Power==""), ]

Chester_Generation_Complete <-  Chester_Generation[!(is.na(Chester_Generation$Power) | Chester_Generation$Power==""), ]

Tri_County_Generation_Complete <-  Tri_County_Generation[!(is.na(Tri_County_Generation$Power) | Tri_County_Generation$Power==""), ]

Glen_Canyon_Generation_Complete <-  Glen_Canyon_Generation[!(is.na(Glen_Canyon_Generation$Power) | Glen_Canyon_Generation$Power==""), ]

Quail_Creek_Generation_Complete <-  Quail_Creek_Generation[!(is.na(Quail_Creek_Generation$Power) | Quail_Creek_Generation$Power==""), ]


```

## Discharge Visualization

```{r}

plot1<-ggplot(gage_01054500, aes(x=date, y=average_monthly_discharge))+geom_line(color="blue")+labs(x="", y="", title = "Gage 01054500")+ylim(0,15000)
plot2<-ggplot(gage_01563200, aes(x=date, y=average_monthly_discharge))+geom_line(color="blue")+labs(x="", y="", title = "Gage 01563200")+ylim(0,15000)
plot3<-ggplot(gage_02071000, aes(x=date, y=average_monthly_discharge))+geom_line(color="blue")+labs(x="", y="", title = "Gage 02071000")+ylim(0,15000)
plot4<-ggplot(gage_02143000, aes(x=date, y=average_monthly_discharge))+geom_line(color="blue")+labs(x="", y="Average Monthly Discharge", title = "Gage 02143000")+ylim(0,15000)
plot5<-ggplot(gage_02163001, aes(x=date, y=average_monthly_discharge))+geom_line(color="blue")+labs(x="", y="", title = "Gage 02163001")+ylim(0,15000)
plot6<-ggplot(gage_06279940, aes(x=date, y=average_monthly_discharge))+geom_line(color="red")+labs(x="", y="", title = "Gage 06279940")+ylim(0,15000)
plot7<-ggplot(gage_09146200, aes(x=date, y=average_monthly_discharge))+geom_line(color="red")+labs(x="", y="", title = "Gage 09146200")+ylim(0,15000)
#plot8<-ggplot(gage_09328990, aes(x=date, y=average_monthly_discharge))+geom_line()
plot9<-ggplot(gage_09408135, aes(x=date, y=average_monthly_discharge))+geom_line(color="red")+labs( x="", y="", title = "Gage 09408135")+ylim(0,15000)+scale_x_date(limits=as.Date(c("2000-01-01", "2021-12-31")))
plot10<-ggplot(gage_13046000, aes(x=date, y=average_monthly_discharge))+geom_line(color="red")+labs(x="", y="", title = "Gage 13046000")+ylim(0,15000)

discharge_plot<-plot_grid(plot1, plot2, plot3, plot4, plot5, plot6, plot7, plot9, plot10, ncol = 3, align="hv")

title_gg <- ggplot() + labs(title = "Discharge Measurements for Streams Associated with Hydropower Generation", subtitle = "Blue indicates eastern states while red indicates western states.")

plot_grid(title_gg, discharge_plot, ncol = 1, rel_heights = c(0.08,1))

```

## Data Analysis

Testing for trends in discharge and generation data.

```{r}
gage_01054500_ts<-ts(gage_01054500[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_01054500<-stl(gage_01054500_ts, s.window = "periodic")
plot(decom_01054500)
trend_gage_01054500<-smk.test(gage_01054500_ts)
trend_gage_01054500
summary(trend_gage_01054500)


gage_01563200_ts<-ts(gage_01563200[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_01563200<- stl(gage_01563200_ts, s.window = "periodic")
plot(decom_01563200)
trend_gage_01563200<-smk.test(gage_01563200_ts)
trend_gage_01563200
summary(trend_gage_01563200)


gage_02071000_ts<-ts(gage_02071000[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_02071000<- stl(gage_02071000_ts, s.window = "periodic")
plot(decom_02071000)
trend_gage_02071000<-smk.test(gage_02071000_ts)
trend_gage_02071000
summary(trend_gage_02071000)


gage_02143000_ts<-ts(gage_02143000[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_02143000<- stl(gage_02143000_ts, s.window = "periodic")
plot(decom_02143000)
trend_gage_02143000<-smk.test(gage_02143000_ts)
trend_gage_02143000
summary(trend_gage_02143000)



gage_02163001_ts<-ts(gage_02163001[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_02163001<- stl(gage_02163001_ts, s.window = "periodic")
plot(decom_02163001)
trend_gage_02163001<-smk.test(gage_02163001_ts)
trend_gage_02163001
summary(trend_gage_02163001)



gage_06279940_ts<-ts(gage_06279940[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_06279940<- stl(gage_06279940_ts, s.window = "periodic")
plot(decom_06279940)
trend_gage_06279940<-smk.test(gage_06279940_ts)
trend_gage_06279940
summary(trend_gage_06279940)



gage_09146200_ts<-ts(gage_09146200[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_09146200<- stl(gage_09146200_ts, s.window = "periodic")
plot(decom_09146200)
trend_gage_09146200<-smk.test(gage_09146200_ts)
trend_gage_09146200
summary(trend_gage_09146200)

#gage_09328990_ts<-ts(gage_09328990[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
#decom_09328990<- stl(gage_09328990_ts, s.window = "periodic")
#plot(decom_09328990)


gage_09408135_ts<-ts(gage_09408135[[3]], frequency = 12,  start = c(2005, 1), end = c(2021, 12))
decom_09408135<- stl(gage_09408135_ts, s.window = "periodic")
plot(decom_09408135)
trend_gage_09408135<-smk.test(gage_09408135_ts)
trend_gage_09408135
summary(trend_gage_09408135)


gage_13046000_ts<-ts(gage_13046000[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_13046000<- stl(gage_13046000_ts, s.window = "periodic")
plot(decom_13046000)
trend_gage_13046000<-smk.test(gage_13046000_ts)
trend_gage_13046000
summary(trend_gage_13046000)


```

```{r}
f_month_1<- month(first(High_Shoals_Generation_Complete$Date))
f_year_1<- year(first((High_Shoals_Generation_Complete$Date)))
high_shoals_ts<- ts(High_Shoals_Generation_Complete$Power,
                   start=c(f_year_1,f_month_1),
                   frequency=12)

f_month_2<- month(first(Schoolfield_Generation_Complete$Date))
f_year_2<- year(first((Schoolfield_Generation_Complete$Date)))
schoolfield_ts<- ts(Schoolfield_Generation_Complete$Power,
                   start=c(f_year_2,f_month_2),
                   frequency=12)

f_month_3<- month(first(Matson_Generation_Complete$Date))
f_year_3<- year(first((Matson_Generation_Complete$Date)))
matson_ts<- ts(Matson_Generation_Complete$Power,
                   start=c(f_year_3,f_month_3),
                   frequency=12)

f_month_4<- month(first(Otis_Generation_Complete$Date))
f_year_4<- year(first((Otis_Generation_Complete$Date)))
otis_ts<- ts(Otis_Generation_Complete$Power,
                   start=c(f_year_4,f_month_4),
                   frequency=12)

f_month_5<- month(first(Ware_Shoals_Generation_Complete$Date))
f_year_5<- year(first((Ware_Shoals_Generation_Complete$Date)))
ware_shoals_ts<- ts(Ware_Shoals_Generation_Complete$Power,
                   start=c(f_year_5,f_month_5),
                   frequency=12)

f_month_6<- month(first(Garland_Generation_Complete$Date))
f_year_6<- year(first((Garland_Generation_Complete$Date)))
garland_ts<- ts(Garland_Generation_Complete$Power,
                   start=c(f_year_6,f_month_6),
                   frequency=12)

f_month_7<- month(first(Chester_Generation_Complete$Date))
f_year_7<- year(first((Chester_Generation_Complete$Date)))
chester_ts<- ts(Chester_Generation_Complete$Power,
                   start=c(f_year_7,f_month_7),
                   frequency=12)


f_month_8<- month(first(Tri_County_Generation_Complete$Date))
f_year_8<- year(first((Tri_County_Generation_Complete$Date)))
tri_county_ts<- ts(Tri_County_Generation_Complete$Power,
                   start=c(f_year_8,f_month_8),
                   frequency=12)

f_month_9<- month(first(Glen_Canyon_Generation_Complete$Date))
f_year_9<- year(first((Glen_Canyon_Generation_Complete$Date)))
glen_canyon_ts<- ts(Glen_Canyon_Generation_Complete$Power,
                   start=c(f_year_9,f_month_9),
                   frequency=12)

f_month_10<- month(first(Quail_Creek_Generation_Complete$Date))
f_year_10<- year(first((Quail_Creek_Generation_Complete$Date)))
quail_creek_ts<- ts(Quail_Creek_Generation_Complete$Power,
                   start=c(f_year_10,f_month_10),
                   frequency=12)

High_Shoals_Generation_Decomposed <- stl(high_shoals_ts, s.window = "periodic")
Schoolfield_Generation_Decomposed <- stl(schoolfield_ts, s.window = "periodic")
Matson_Generation_Decomposed <- stl(matson_ts, s.window = "periodic")
Otis_Generation_Decomposed <- stl(otis_ts, s.window = "periodic")
Ware_Shoals_Generation_Decomposed <- stl(ware_shoals_ts, s.window = "periodic")
Garland_Generation_Decomposed <- stl(garland_ts, s.window = "periodic")
Chester_Generation_Decomposed <- stl(chester_ts, s.window = "periodic")
Tri_County_Generation_Decomposed <- stl(tri_county_ts, s.window = "periodic")
Glen_Canyon_Generation_Decomposed <- stl(glen_canyon_ts, s.window = "periodic")
Quail_Creek_Generation_Decomposed <- stl(quail_creek_ts, s.window = "periodic")

High_Shoals_Generation_Components <-as.data.frame(High_Shoals_Generation_Decomposed$time.series[,1:3])
Schoolfield_Generation_Components <- as.data.frame(Schoolfield_Generation_Decomposed$time.series[,1:3])
Matson_Generation_Components <- as.data.frame(Matson_Generation_Decomposed$time.series[,1:3])
Otis_Generation_Components <- as.data.frame(Otis_Generation_Decomposed$time.series[,1:3])
Ware_Shoals_Generation_Components <- as.data.frame(Ware_Shoals_Generation_Decomposed$time.series[,1:3])
Garland_Generation_Components <- as.data.frame(Garland_Generation_Decomposed$time.series[,1:3])
Chester_Generation_Components <- as.data.frame(Chester_Generation_Decomposed$time.series[,1:3])
Tri_County_Generation_Components <- as.data.frame(Tri_County_Generation_Decomposed$time.series[,1:3])
Glen_Canyon_Generation_Components <- as.data.frame(Glen_Canyon_Generation_Decomposed$time.series[,1:3])
Quail_Creek_Generation_Components <- as.data.frame(Quail_Creek_Generation_Decomposed$time.series[,1:3])

High_Shoals_Generation_Components <- mutate(High_Shoals_Generation_Components,
        Observed = High_Shoals_Generation_Complete$Power,     
        Date = High_Shoals_Generation_Complete$Date)


Schoolfield_Generation_Components <- mutate(Schoolfield_Generation_Components,
        Observed = Schoolfield_Generation_Complete$Power,     
        Date = Schoolfield_Generation_Complete$Date)


Matson_Generation_Components <- mutate(Matson_Generation_Components,
        Observed = Matson_Generation_Complete$Power,     
        Date = Matson_Generation_Complete$Date)


Otis_Generation_Components <- mutate(Otis_Generation_Components,
        Observed = Otis_Generation_Complete$Power,     
        Date = Otis_Generation_Complete$Date)


Ware_Shoals_Generation_Components <- mutate(Ware_Shoals_Generation_Components,
        Observed = Ware_Shoals_Generation_Complete$Power,     
        Date = Ware_Shoals_Generation_Complete$Date)

Garland_Generation_Components <- mutate(Garland_Generation_Components,
        Observed = Garland_Generation_Complete$Power,     
        Date = Garland_Generation_Complete$Date)


Chester_Generation_Components <- mutate(Chester_Generation_Components,
        Observed = Chester_Generation_Complete$Power,     
        Date = Chester_Generation_Complete$Date)


Tri_County_Generation_Components <- mutate(Tri_County_Generation_Components,
        Observed = Tri_County_Generation_Complete$Power,     
        Date = Tri_County_Generation_Complete$Date)


Glen_Canyon_Generation_Components <- mutate(Glen_Canyon_Generation_Components,
        Observed = Glen_Canyon_Generation_Complete$Power,     
        Date = Glen_Canyon_Generation_Complete$Date)

Quail_Creek_Generation_Components <- mutate(Quail_Creek_Generation_Components,
        Observed = Quail_Creek_Generation_Complete$Power,     
        Date = Quail_Creek_Generation_Complete$Date)

# Run SMK test
high_shoals_trend1 <- Kendall::SeasonalMannKendall(high_shoals_ts)
# Inspect results
high_shoals_trend1
summary(high_shoals_trend1)

# Run SMK test
schoolfield_trend1 <- Kendall::SeasonalMannKendall(schoolfield_ts)
# Inspect results
schoolfield_trend1
summary(schoolfield_trend1)

# Run SMK test
matson_trend1 <- Kendall::SeasonalMannKendall(matson_ts)
# Inspect results
matson_trend1
summary(matson_trend1)

# Run SMK test
otis_trend1 <- Kendall::SeasonalMannKendall(otis_ts)
# Inspect results
otis_trend1
summary(otis_trend1)

# Run SMK test
ware_shoals_trend1 <- Kendall::SeasonalMannKendall(ware_shoals_ts)
# Inspect results
ware_shoals_trend1
summary(ware_shoals_trend1)


# Run SMK test
garland_trend1 <- Kendall::SeasonalMannKendall(garland_ts)
# Inspect results
garland_trend1
summary(garland_trend1)


# Run SMK test
chester_trend1 <- Kendall::SeasonalMannKendall(chester_ts)
# Inspect results
chester_trend1
summary(chester_trend1)


# Run SMK test
tri_county_trend1 <- Kendall::SeasonalMannKendall(tri_county_ts)
# Inspect results
tri_county_trend1
summary(tri_county_trend1)

# Run SMK test
glen_canyon_trend1 <- Kendall::SeasonalMannKendall(glen_canyon_ts)
# Inspect results
glen_canyon_trend1
summary(glen_canyon_trend1)


# Run SMK test
quail_creek_trend1 <- Kendall::SeasonalMannKendall(quail_creek_ts)
# Inspect results
quail_creek_trend1
summary(quail_creek_trend1)


```

```{r}
# Visualize how the trend maps onto the data

ggplot(High_Shoals_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Schoolfield_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Matson_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Otis_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Ware_Shoals_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Garland_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Chester_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Tri_County_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Glen_Canyon_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Quail_Creek_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

# Visualize how the seasonal cycle maps onto the data
ggplot(High_Shoals_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Schoolfield_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Matson_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Otis_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Ware_Shoals_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Garland_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Chester_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Tri_County_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Glen_Canyon_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Quail_Creek_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

```

## Combining Datasets 

```{r}
#Matching the stream gage data to the power generation data.

#High Shoals Dam
high_shoals_power_water<-left_join(High_Shoals_Generation_Complete, gage_02143000)%>%
  mutate(Power=as.numeric(Power))

ggplot(high_shoals_power_water, aes(x=average_monthly_discharge, y=Power))+geom_point()+scale_y_log10()+geom_smooth()


schoolfield_power_water<-left_join(Schoolfield_Generation_Complete, gage_02071000)%>%
  mutate(Power=as.numeric(Power))

ggplot(schoolfield_power_water, aes(x=average_monthly_discharge, y=Power))+geom_point()


garland_power_water<-left_join(Garland_Generation_Complete, gage_06279940)%>%
  mutate(Power=as.numeric(Power))

ggplot(garland_power_water, aes(x=average_monthly_discharge, y=Power))+geom_point()

```

## Results 