---
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---
title: "Final Project - Hydro Power Generation and Streamflow Analysis"
author: "Kaitlyn Elliott and Jon Nichols"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Rationale and Research Questions

Hydroelectric power accounts for 52 percent of renewable electricity generation and 7 percent of total electricity generation in the US. These plants are a critical clean energy resource for grid operators and utilities due to their operational flexibility, low maintenance costs, and non-intermittent production. However, hydroelectric generation is susceptible to climate change, and changes in temperature, precipitation patterns, glacial melt, and the frequency of extreme weather events such as floods and droughts may have a direct negative effect on electricity production. Thus, the impact of climate change on the variability and availability of streamflow is critical for current hydroelectric generation and future operational forecast plans. 

There are three types of hydropower plants: impoundment facilities that use a dam to store water, run-of-river facilities that generate electricity by channeling water through a turbine and do not use a reservoir, and pumped storage plants. For our analysis, we investigated the impact of climate change on both impoundment facilities and run-of-river facilities. We expect that run-of-river facilities will experience more significant declines in generation given their unique vulnerability to variations in streamflow. In general, lower stream discharges reduce the amount of electricity a hydropower plant can generate. 

We examine 20 years of streamflow and electric generation data for 9 facilities across the US and assess the impact of climate change on stream flows and consequently on the hydropower generation of these plants. We specifically investigate the monotonic trends for streamflow and electric generation at each hydropower plant, and then analyze whether stream discharge can explain the variation in observed generation data. The results obtained will allow us to better understand the relationship between hydropower generation and streamflow, while also identifying hydropower plants that are vulnerable to climate change.


## Background and Methods
In this project we set out to explore the relationship between hydro power generation and stream generation. To do this we looked at power generation data from the EPI and stream gage data from the USGS. For generation data we selected generation sites/dams so that half were from the western United States and half from the East. Stream gage data was selected upstream from dams and reservoirs to get an accurate representation of stream flow over time. Both power generation data and stream flow was analyzed for seasonality and monotonic trends using a seasonal man kendall test. This was to help understand if generation was increasing or decreasing as a result of stream flow. Although outside factors could also contribute increases or decreases in power generation.

## Set Up  

```{r}

getwd()

library(tidyverse)
library(lubridate)
library(dataRetrieval)
library(trend)
library(dataRetrieval)
library(cowplot)
library(readr)
library(zoo)
library(Kendall)
#install.packages("tseries")
library(tseries)
library(grid)
library(gridExtra)
library(knitr)
library(sf)
library(mapview)
library(maps)

theme_set(theme_classic())
options(scipen = 6)

```

## Data Retrieval

Hydroelectric generation data was collected from the EIA, and streamflow data was collected from the USGS. Generation data was obtained through the EIA's Electricity Data Browser, which is an interactive query tool. Nine separate queries - one for each hydropower plant- were used to download the data needed for the analysis. For each query, we used the date range January 2001 through December 2021. Data was downloaded as a "Chart (CSV)", which presented four rows of descriptive information and then two columns for the Month and "All primemovers - All fuels (ALL) megawatthours". The descriptive info of each generation dataset was cleaned in Excel to streamline data import into R. 

Detail   | Description
---------| ------------- 
Data Source | EIA Net Generation Data   
Retrieved From | https://www.eia.gov/beta/electricity/data/browser/#/plant/10550/?freq=M&pin=
Variables Used | Month, "All fuels (ALL) megawatthours"
Date Range | January 2001 - December 2021

NEED HELP WITH CREATING A TITLE FOR THAT TABLE 
SOMETHING ABOUT RETRIEVING USGS DATA 


## Data Wrangling

After streamflow data was imported, the column containing discharge data was mutated and titled “Discharge”. Next, three columns were selected: Discharge, Datetime, and Site Number. The Datetime was mutated to produce a Year column and a Month column. We then grouped the dataset by the year and month, and created an “Average Monthly Discharge” column. This process was repeated for each stream gauge dataset. Additionally, an arbitrary day - the first of each month - was added to each imported dataset to enable proper formatting. 

After hydroelectric generation data was imported, the Date column had to be wrangled and reformatted to allow for further analysis. Month and year columns were added to each dataset, generation data was transformed to be numeric, the date column was converted to a date variable in R, and data was arranged in ascending fashion. Empty or “NA” values were then dropped to have complete sets of data for each hydroelectric plant. 

After importing and wrangling hydroelectric generation and streamflow data separately, we used a left join function to combine the datasets for each specific facility. Given that some generation data was missing for the 20-year period, we used left join to match the streamflow data to the generation data that did exist for each site. 




```{r}
#Reading in stream data

#East Stream Gage Sites

gage_02143000<-readNWISdata(siteNumbers = "02143000", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_02071000<-readNWISdata(siteNumbers = "02071000", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_01563200<-readNWISdata(siteNumbers = "01563200", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_01054500<-readNWISdata(siteNumbers = "01054500", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_02163001<-readNWISdata(siteNumbers = "02163001", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

#West Stream Gage Sites

gage_06279940<-readNWISdata(siteNumbers = "06279940", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_13046000<-readNWISdata(siteNumbers = "13046000", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_09146200<-readNWISdata(siteNumbers = "09146200", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

gage_09408135<-readNWISdata(siteNumbers = "09408135", parameterCd = "00060", startDate = "2000-01-01", endDate = "2021-12-31")

#Summarizing Data by Month

gage_01054500 <- gage_01054500 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_01563200<-gage_01563200 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_02071000<-gage_02071000 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_02143000<-gage_02143000 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_02163001<-gage_02163001 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_06279940<-gage_06279940 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_09146200<-gage_09146200 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))


gage_09408135<-gage_09408135 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))

gage_13046000<-gage_13046000 %>%
  mutate(Discharge = X_00060_00003) %>%
  select(Discharge, dateTime, site_no) %>%
  mutate(Month = month(dateTime), Year= year(dateTime)) %>%
  group_by(Year, Month) %>%
  summarize(average_monthly_discharge = mean(Discharge)) %>%
  mutate(day="01",date=as.Date(paste(Month, day, Year, sep = "-"),
                               format="%m-%d-%Y"))
#Reading in generation data
High_Shoals_Generation <- read.csv("./Final Project Hydro Generation Data/High_Shoals_Hydro_(NC)_net_generation (1).csv")

Schoolfield_Generation <- read.csv("./Final Project Hydro Generation Data/Schoolfield_Dam_net_generation.csv")

Matson_Generation <- read.csv("./Final Project Hydro Generation Data/William_F_Matson_Generating_Station_net_generation.csv")

Otis_Generation <- read.csv("./Final Project Hydro Generation Data/Otis_Hydro_net_generation.csv")

Ware_Shoals_Generation <- read.csv("./Final Project Hydro Generation Data/Ware_Shoals_Hydro_Project_net_generation.csv")

Garland_Generation <- read.csv("./Final Project Hydro Generation Data/Garland_Canal_Power_Plant_net_generation.csv")

Chester_Generation <- read.csv("./Final Project Hydro Generation Data/Chester_Diversion_Hydroelectric_Project_net_generation.csv")

Tri_County_Generation <- read.csv("./Final Project Hydro Generation Data/Tri-County_Water_Hydropower_Project_net_generation.csv")

Quail_Creek_Generation <- read.csv("./Final Project Hydro Generation Data/Quail_Creek_Hydro_Plant_#1_net_generation.csv")


High_Shoals_Generation <- High_Shoals_Generation[-c(1),]
Schoolfield_Generation <- Schoolfield_Generation[-c(1),]
Matson_Generation <- Matson_Generation[-c(1),]
Otis_Generation <- Otis_Generation[-c(1),]
Ware_Shoals_Generation <-Ware_Shoals_Generation[-c(1),] 
Garland_Generation <- Garland_Generation[-c(1),]
Chester_Generation <- Chester_Generation[-c(1),]
Tri_County_Generation <- Tri_County_Generation[-c(1),]
Quail_Creek_Generation <- Quail_Creek_Generation[-c(1),]

colnames(High_Shoals_Generation) <- c("date_wrong", "Power")
colnames(Schoolfield_Generation) <- c("date_wrong", "Power")
colnames(Matson_Generation) <- c("date_wrong", "Power")
colnames(Otis_Generation) <- c("date_wrong", "Power")
colnames(Ware_Shoals_Generation) <- c("date_wrong", "Power")
colnames(Garland_Generation) <- c("date_wrong", "Power")
colnames(Chester_Generation) <- c("date_wrong", "Power")
colnames(Tri_County_Generation) <- c("date_wrong", "Power")
colnames(Quail_Creek_Generation) <- c("date_wrong", "Power")

High_Shoals_Generation <- High_Shoals_Generation %>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Schoolfield_Generation <- Schoolfield_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Matson_Generation <- Matson_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Otis_Generation <- Otis_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Ware_Shoals_Generation <-Ware_Shoals_Generation %>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Garland_Generation <- Garland_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Chester_Generation <- Chester_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Tri_County_Generation <- Tri_County_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))
Quail_Creek_Generation <- Quail_Creek_Generation%>%
  mutate(Date = as.Date(paste(date_wrong, "-01", sep=""), format = "%b-%y-%d"))

High_Shoals_Generation <- High_Shoals_Generation %>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)
  

Schoolfield_Generation <- Schoolfield_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Matson_Generation <- Matson_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Otis_Generation <- Otis_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Ware_Shoals_Generation <-Ware_Shoals_Generation %>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Garland_Generation <- Garland_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Chester_Generation <- Chester_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)

Tri_County_Generation <- Tri_County_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)


Quail_Creek_Generation <- Quail_Creek_Generation%>%
  mutate(Month = month(Date))%>%
mutate(Year = year(Date))%>%
  select(Date, Month, Year, Power)%>%
  transform(Power = as.numeric(Power))%>%
  arrange(Date)



High_Shoals_Generation_Complete <-  High_Shoals_Generation[!(is.na(High_Shoals_Generation$Power) | High_Shoals_Generation$Power==""), ]

Schoolfield_Generation_Complete <-  Schoolfield_Generation[!(is.na(Schoolfield_Generation$Power) | Schoolfield_Generation$Power==""), ]

Matson_Generation_Complete <-  Matson_Generation[!(is.na(Matson_Generation$Power) | Matson_Generation$Power==""), ]

Otis_Generation_Complete <-  Otis_Generation[!(is.na(Otis_Generation$Power) | Otis_Generation$Power==""), ]

Ware_Shoals_Generation_Complete <-  Ware_Shoals_Generation[!(is.na(Ware_Shoals_Generation$Power) | Ware_Shoals_Generation$Power==""), ]

Garland_Generation_Complete <-  Garland_Generation[!(is.na(Garland_Generation$Power) | Garland_Generation$Power==""), ]

Chester_Generation_Complete <-  Chester_Generation[!(is.na(Chester_Generation$Power) | Chester_Generation$Power==""), ]

Tri_County_Generation_Complete <-  Tri_County_Generation[!(is.na(Tri_County_Generation$Power) | Tri_County_Generation$Power==""), ]

Quail_Creek_Generation_Complete <-  Quail_Creek_Generation[!(is.na(Quail_Creek_Generation$Power) | Quail_Creek_Generation$Power==""), ]


```

## Discharge Visualization

```{r}

plot1<-ggplot(gage_01054500, aes(x=date, y=average_monthly_discharge))+geom_line(color="blue")+labs(x="", y="", title = "Gage 01054500")+ylim(0,15000)
plot2<-ggplot(gage_01563200, aes(x=date, y=average_monthly_discharge))+geom_line(color="blue")+labs(x="", y="", title = "Gage 01563200")+ylim(0,15000)
plot3<-ggplot(gage_02071000, aes(x=date, y=average_monthly_discharge))+geom_line(color="blue")+labs(x="", y="", title = "Gage 02071000")+ylim(0,15000)
plot4<-ggplot(gage_02143000, aes(x=date, y=average_monthly_discharge))+geom_line(color="blue")+labs(x="", y="", title = "Gage 02143000")+ylim(0,15000)
plot5<-ggplot(gage_02163001, aes(x=date, y=average_monthly_discharge))+geom_line(color="blue")+labs(x="", y="", title = "Gage 02163001")+ylim(0,15000)
plot6<-ggplot(gage_06279940, aes(x=date, y=average_monthly_discharge))+geom_line(color="red")+labs(x="", y="", title = "Gage 06279940")+ylim(0,15000)
plot7<-ggplot(gage_09146200, aes(x=date, y=average_monthly_discharge))+geom_line(color="red")+labs(x="", y="", title = "Gage 09146200")+ylim(0,15000)
plot9<-ggplot(gage_09408135, aes(x=date, y=average_monthly_discharge))+geom_line(color="red")+labs( x="", y="", title = "Gage 09408135")+ylim(0,15000)+scale_x_date(limits=as.Date(c("2000-01-01", "2021-12-31")))
plot10<-ggplot(gage_13046000, aes(x=date, y=average_monthly_discharge))+geom_line(color="red")+labs(x="", y="", title = "Gage 13046000")+ylim(0,15000)

discharge_plot<-plot_grid(plot1, plot2, plot3, plot4, plot5, plot6, plot7, plot9, plot10, ncol = 3, align="hv")

title_gg <- ggplot() + labs(title = "Figure 1: Discharge Measurements for Streams Associated with Hydropower Generation", subtitle = "Blue indicates eastern states while red indicates western states.")

discharge_plot_grid<-plot_grid(title_gg, discharge_plot, ncol = 1, rel_heights = c(0.08,1))+theme(plot.margin=margin(10,5,10,10))

y.grob_dis<-textGrob(~"Average Monthly Discharge "(ft^3/s), 
                   gp=gpar(col="black", fontsize=10), rot=90)

grid.arrange(arrangeGrob(discharge_plot_grid, left = y.grob_dis))

```


## Hydroelectric Generation Visualization 

```{r}
#cowplots for hydro generation 

plot_hs<-ggplot(High_Shoals_Generation, aes(x=Date, y=Power))+geom_line(color="blue")+labs(x="", y="", title = "High Shoals")+ylim(0,4000)

plot_school<-ggplot(Schoolfield_Generation, aes(x=Date, y=Power))+geom_line(color="blue")+labs(x="", y="", title = "Schoolfield")+ylim(0,4000)

plot_matson<-ggplot(Matson_Generation, aes(x=Date, y=Power))+geom_line(color="blue")+labs(x="", y="", title = "Matson")+ylim(0,16000)

plot_otis<-ggplot(Otis_Generation, aes(x=Date, y=Power))+geom_line(color="blue")+labs(x="", y="", title = "Otis")+ylim(0,10000)

plot_ws<-ggplot(Ware_Shoals_Generation, aes(x=Date, y=Power))+geom_line(color="blue")+labs(x="", y="", title = "Ware Shoals")+ylim(0,7500)

plot_gar<-ggplot(Garland_Generation, aes(x=Date, y=Power))+geom_line(color="red")+labs(x="", y="", title = "Garland")+ylim(0,4000)

plot_ches<-ggplot(Chester_Generation, aes(x=Date, y=Power))+geom_line(color="red")+labs(x="", y="", title = "Chester")+ylim(0,2000)

plot_tri<-ggplot(Tri_County_Generation, aes(x=Date, y=Power))+geom_line(color="red")+labs(x="", y="", title = "Tri County")+ylim(0,7500)


plot_quail<-ggplot(Quail_Creek_Generation, aes(x=Date, y=Power))+geom_line(color="red")+labs(x="", y="", title = "Quail Creek")+ylim(0,3000)

power_plot<-plot_grid(plot_hs, plot_school, plot_matson, plot_otis, plot_ws, plot_gar, plot_ches, plot_tri,plot_quail, ncol = 3, align="hv")

title_gg_2 <- ggplot() + labs(title = "Figure 2: Hydropower Generation for Sampled Power Plants", subtitle = "Blue indicates eastern states while red indicates western states.")

plot_grid_power<-plot_grid(title_gg_2, power_plot, ncol = 1, rel_heights = c(0.08,1))+theme(plot.margin=margin(30, 10, 10, 10))

y.grob_power<-textGrob("Power Generation (MWh)", 
                   gp=gpar(col="black", fontsize=10), rot=90)

grid.arrange(arrangeGrob(plot_grid_power, left = y.grob_power))



```


## Data Analysis

Question 1: Are there trends in the streamflow data? 

Following our initial wrangling, we created plot grids to visualize the streamflow and generation data. Figure 1 displays the discharge measurements for the streams associated with the selected hydropower plants. Figure 2 displays the electric generation for the sampled power plants. Both visualizations compared sites in the Western US to sites in the Eastern US. Both graphs appeared to show strong seasonal components. 

After visualizing the data, we investigated trends in streamflow. We created a time series for each stream gauge, and then decomposed each time series into its seasonal, trend, and remainder components. SOMETHING ABOUT RESULT

We ran a Seasonal Mann-Kendall test on each time series to test whether there was a monotonic trend in streamflow over time. The results are displayed in Table 1. 



```{r}
gage_01054500_ts<-ts(gage_01054500[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_01054500<-stl(gage_01054500_ts, s.window = "periodic")
plot(decom_01054500)
trend_gage_01054500<-smk.test(gage_01054500_ts)
trend_gage_01054500
summary(trend_gage_01054500)


gage_01563200_ts<-ts(gage_01563200[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_01563200<- stl(gage_01563200_ts, s.window = "periodic")
plot(decom_01563200)
trend_gage_01563200<-smk.test(gage_01563200_ts)
trend_gage_01563200
summary(trend_gage_01563200)


gage_02071000_ts<-ts(gage_02071000[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_02071000<- stl(gage_02071000_ts, s.window = "periodic")
plot(decom_02071000)
trend_gage_02071000<-smk.test(gage_02071000_ts)
trend_gage_02071000
summary(trend_gage_02071000)


gage_02143000_ts<-ts(gage_02143000[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_02143000<- stl(gage_02143000_ts, s.window = "periodic")
plot(decom_02143000)
trend_gage_02143000<-smk.test(gage_02143000_ts)
trend_gage_02143000
summary(trend_gage_02143000)



gage_02163001_ts<-ts(gage_02163001[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_02163001<- stl(gage_02163001_ts, s.window = "periodic")
plot(decom_02163001)
trend_gage_02163001<-smk.test(gage_02163001_ts)
trend_gage_02163001
summary(trend_gage_02163001)



gage_06279940_ts<-ts(gage_06279940[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_06279940<- stl(gage_06279940_ts, s.window = "periodic")
plot(decom_06279940)
trend_gage_06279940<-smk.test(gage_06279940_ts)
trend_gage_06279940
summary(trend_gage_06279940)



gage_09146200_ts<-ts(gage_09146200[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_09146200<- stl(gage_09146200_ts, s.window = "periodic")
plot(decom_09146200)
trend_gage_09146200<-smk.test(gage_09146200_ts)
trend_gage_09146200
summary(trend_gage_09146200)


gage_09408135_ts<-ts(gage_09408135[[3]], frequency = 12,  start = c(2005, 1), end = c(2021, 12))
decom_09408135<- stl(gage_09408135_ts, s.window = "periodic")
plot(decom_09408135)
trend_gage_09408135<-smk.test(gage_09408135_ts)
trend_gage_09408135
summary(trend_gage_09408135)


gage_13046000_ts<-ts(gage_13046000[[3]], frequency = 12,  start = c(2000, 1), end = c(2021, 12))
decom_13046000<- stl(gage_13046000_ts, s.window = "periodic")
plot(decom_13046000)
trend_gage_13046000<-smk.test(gage_13046000_ts)
trend_gage_13046000
summary(trend_gage_13046000)
#kable(trend_gage_13046000)


```

Question 2: Are there trends in the generation data? 


Additionally, we ran a Seasonal Mann-Kendall test on each time series to test whether there was a monotonic trend in hydroelectric generation over time. The results are displayed in Table 2. Two plants, High Shoals and Quail Creek, display a slight statististically significant positive trend over time. Conversely, Ware Shoals and Tri County display a statistically significant negative trend over time. Tri County, which is located in Colorado, has a relatively substantial negative trend. The other hydropower plants have 2-sided P-values greater than 0.05, and thus we cannot reject the null hypothesis that no trend exists. Overall, the results indicate that hydropower generation demonstrates no clear trend on the national level, and each power plant is subject to unique conditions that may cause generation to trend in either direction. 


```{r}
f_month_1<- month(first(High_Shoals_Generation_Complete$Date))
f_year_1<- year(first((High_Shoals_Generation_Complete$Date)))
high_shoals_ts<- ts(High_Shoals_Generation_Complete$Power,
                   start=c(f_year_1,f_month_1),
                   frequency=12)

f_month_2<- month(first(Schoolfield_Generation_Complete$Date))
f_year_2<- year(first((Schoolfield_Generation_Complete$Date)))
schoolfield_ts<- ts(Schoolfield_Generation_Complete$Power,
                   start=c(f_year_2,f_month_2),
                   frequency=12)

f_month_3<- month(first(Matson_Generation_Complete$Date))
f_year_3<- year(first((Matson_Generation_Complete$Date)))
matson_ts<- ts(Matson_Generation_Complete$Power,
                   start=c(f_year_3,f_month_3),
                   frequency=12)

f_month_4<- month(first(Otis_Generation_Complete$Date))
f_year_4<- year(first((Otis_Generation_Complete$Date)))
otis_ts<- ts(Otis_Generation_Complete$Power,
                   start=c(f_year_4,f_month_4),
                   frequency=12)

f_month_5<- month(first(Ware_Shoals_Generation_Complete$Date))
f_year_5<- year(first((Ware_Shoals_Generation_Complete$Date)))
ware_shoals_ts<- ts(Ware_Shoals_Generation_Complete$Power,
                   start=c(f_year_5,f_month_5),
                   frequency=12)

f_month_6<- month(first(Garland_Generation_Complete$Date))
f_year_6<- year(first((Garland_Generation_Complete$Date)))
garland_ts<- ts(Garland_Generation_Complete$Power,
                   start=c(f_year_6,f_month_6),
                   frequency=12)

f_month_7<- month(first(Chester_Generation_Complete$Date))
f_year_7<- year(first((Chester_Generation_Complete$Date)))
chester_ts<- ts(Chester_Generation_Complete$Power,
                   start=c(f_year_7,f_month_7),
                   frequency=12)


f_month_8<- month(first(Tri_County_Generation_Complete$Date))
f_year_8<- year(first((Tri_County_Generation_Complete$Date)))
tri_county_ts<- ts(Tri_County_Generation_Complete$Power,
                   start=c(f_year_8,f_month_8),
                   frequency=12)


f_month_10<- month(first(Quail_Creek_Generation_Complete$Date))
f_year_10<- year(first((Quail_Creek_Generation_Complete$Date)))
quail_creek_ts<- ts(Quail_Creek_Generation_Complete$Power,
                   start=c(f_year_10,f_month_10),
                   frequency=12)

High_Shoals_Generation_Decomposed <- stl(high_shoals_ts, s.window = "periodic")
Schoolfield_Generation_Decomposed <- stl(schoolfield_ts, s.window = "periodic")
Matson_Generation_Decomposed <- stl(matson_ts, s.window = "periodic")
Otis_Generation_Decomposed <- stl(otis_ts, s.window = "periodic")
Ware_Shoals_Generation_Decomposed <- stl(ware_shoals_ts, s.window = "periodic")
Garland_Generation_Decomposed <- stl(garland_ts, s.window = "periodic")
Chester_Generation_Decomposed <- stl(chester_ts, s.window = "periodic")
Tri_County_Generation_Decomposed <- stl(tri_county_ts, s.window = "periodic")
Quail_Creek_Generation_Decomposed <- stl(quail_creek_ts, s.window = "periodic")

High_Shoals_Generation_Components <-as.data.frame(High_Shoals_Generation_Decomposed$time.series[,1:3])
Schoolfield_Generation_Components <- as.data.frame(Schoolfield_Generation_Decomposed$time.series[,1:3])
Matson_Generation_Components <- as.data.frame(Matson_Generation_Decomposed$time.series[,1:3])
Otis_Generation_Components <- as.data.frame(Otis_Generation_Decomposed$time.series[,1:3])
Ware_Shoals_Generation_Components <- as.data.frame(Ware_Shoals_Generation_Decomposed$time.series[,1:3])
Garland_Generation_Components <- as.data.frame(Garland_Generation_Decomposed$time.series[,1:3])
Chester_Generation_Components <- as.data.frame(Chester_Generation_Decomposed$time.series[,1:3])
Tri_County_Generation_Components <- as.data.frame(Tri_County_Generation_Decomposed$time.series[,1:3])
Quail_Creek_Generation_Components <- as.data.frame(Quail_Creek_Generation_Decomposed$time.series[,1:3])

High_Shoals_Generation_Components <- mutate(High_Shoals_Generation_Components,
        Observed = High_Shoals_Generation_Complete$Power,     
        Date = High_Shoals_Generation_Complete$Date)


Schoolfield_Generation_Components <- mutate(Schoolfield_Generation_Components,
        Observed = Schoolfield_Generation_Complete$Power,     
        Date = Schoolfield_Generation_Complete$Date)


Matson_Generation_Components <- mutate(Matson_Generation_Components,
        Observed = Matson_Generation_Complete$Power,     
        Date = Matson_Generation_Complete$Date)


Otis_Generation_Components <- mutate(Otis_Generation_Components,
        Observed = Otis_Generation_Complete$Power,     
        Date = Otis_Generation_Complete$Date)


Ware_Shoals_Generation_Components <- mutate(Ware_Shoals_Generation_Components,
        Observed = Ware_Shoals_Generation_Complete$Power,     
        Date = Ware_Shoals_Generation_Complete$Date)

Garland_Generation_Components <- mutate(Garland_Generation_Components,
        Observed = Garland_Generation_Complete$Power,     
        Date = Garland_Generation_Complete$Date)


Chester_Generation_Components <- mutate(Chester_Generation_Components,
        Observed = Chester_Generation_Complete$Power,     
        Date = Chester_Generation_Complete$Date)


Tri_County_Generation_Components <- mutate(Tri_County_Generation_Components,
        Observed = Tri_County_Generation_Complete$Power,     
        Date = Tri_County_Generation_Complete$Date)



Quail_Creek_Generation_Components <- mutate(Quail_Creek_Generation_Components,
        Observed = Quail_Creek_Generation_Complete$Power,     
        Date = Quail_Creek_Generation_Complete$Date)

# Run SMK test
high_shoals_trend1 <- Kendall::SeasonalMannKendall(high_shoals_ts)
# Inspect results
high_shoals_trend1
summary(high_shoals_trend1)

# Run SMK test
schoolfield_trend1 <- Kendall::SeasonalMannKendall(schoolfield_ts)
# Inspect results
schoolfield_trend1
summary(schoolfield_trend1)

# Run SMK test
matson_trend1 <- Kendall::SeasonalMannKendall(matson_ts)
# Inspect results
matson_trend1
summary(matson_trend1)

# Run SMK test
otis_trend1 <- Kendall::SeasonalMannKendall(otis_ts)
# Inspect results
otis_trend1
summary(otis_trend1)

# Run SMK test
ware_shoals_trend1 <- Kendall::SeasonalMannKendall(ware_shoals_ts)
# Inspect results
ware_shoals_trend1
summary(ware_shoals_trend1)


# Run SMK test
garland_trend1 <- Kendall::SeasonalMannKendall(garland_ts)
# Inspect results
garland_trend1
summary(garland_trend1)


# Run SMK test
chester_trend1 <- Kendall::SeasonalMannKendall(chester_ts)
# Inspect results
chester_trend1
summary(chester_trend1)


# Run SMK test
tri_county_trend1 <- Kendall::SeasonalMannKendall(tri_county_ts)
# Inspect results
tri_county_trend1
summary(tri_county_trend1)


# Run SMK test
quail_creek_trend1 <- Kendall::SeasonalMannKendall(quail_creek_ts)
# Inspect results
quail_creek_trend1
summary(quail_creek_trend1)

generation_smk_results <- data.frame(
  Hydropower_Plants = c("High Shoals", "Schoolfield", "Matson", "Otis", "Ware Shoals", "Garland", "Chetser", "Tri County", "Quail Creek"),
  tau = c(0.097,0.026,0.050,-0.052,-0.105,-0.102,0.075,-0.292,0.100),
  Two_sided_P_value = c(0.033,0.562,0.271,0.249,0.021,0.115,0.386,0.001,0.028)
)


generation_smk_results %>%
  kbl(caption = "Seasonal Mann Kendall Tests for Hydroelectric Generation") %>%
  kable_classic(full_width = F, html_font = "Cambria")



```

ggplot was used to graphically represent the trend and seasonal components relative to the observed generation data. Additional figures are included in the R code file associated with this report. ggplot was also used to display scatterplots of streamflow and power generation for each hydroelectric plant analyzed, as seen in Figure 3. 

```{r}
# Visualize how the trend maps onto the data

ggplot(High_Shoals_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ggtitle("High Shoals Trend Analysis")+
  ylab(expression("Power (MWh)"))


ggplot(Schoolfield_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ggtitle("Schoolfield Trend Analysis")+
  ylab(expression("Power (MWh)"))

ggplot(Matson_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Otis_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Ware_Shoals_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Garland_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Chester_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Tri_County_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ggtitle("Tri County Trend Analysis")+
  ylab(expression("Power (MWh)"))


ggplot(Quail_Creek_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = trend, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

# Visualize how the seasonal cycle maps onto the data
ggplot(High_Shoals_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  ggtitle("High Shoals Seasonality Analysis")+
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Schoolfield_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Matson_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Otis_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

ggplot(Ware_Shoals_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Garland_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Chester_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))


ggplot(Tri_County_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))



ggplot(Quail_Creek_Generation_Components) +
  geom_line(aes(y = Observed, x = Date),  size = 0.25) +
  geom_line(aes(y = seasonal, x = Date), color = "#c13d75ff") +
  geom_hline(yintercept = 0, lty = 2) +
  ylab(expression("Power (MWh)"))

```

## Combining Datasets 

```{r}
#Matching the stream gage data to the power generation data.

#High Shoals Dam
high_shoals_power_water<-left_join(High_Shoals_Generation_Complete, gage_02143000)

hs_combined_plot <- ggplot(high_shoals_power_water, aes(x=average_monthly_discharge, y=Power))+geom_point(alpha=0.8, color = "blue")+labs(x=~"", y="", title="High Shoals")




schoolfield_power_water<-left_join(Schoolfield_Generation_Complete, gage_02071000)

schoolfield_combined_plot <- ggplot(schoolfield_power_water, aes(x=average_monthly_discharge, y=Power))+geom_point(alpha=0.8, color = "blue")+labs(x=~"", y="", title="Schoolfield")




william_f_matson_power_water<-left_join(Matson_Generation_Complete, gage_01563200)

matson_combined_plot <- ggplot(william_f_matson_power_water, aes(x=average_monthly_discharge, y=Power))+geom_point(alpha=0.8, color = "blue")+labs(x=~"", y="", title="William F Matson")




otis_power_water<-left_join(Otis_Generation_Complete, gage_01054500)

otis_combined_plot <- ggplot(otis_power_water, aes(x=average_monthly_discharge, y=Power))+geom_point(apha=0.8, color = "blue")+labs(x=~"", y="", title="Otis")




ware_power_water<-left_join(Ware_Shoals_Generation_Complete, gage_02163001)

ws_combined_plot <- ggplot(ware_power_water, aes(x=average_monthly_discharge, y=Power))+geom_point(alpha=0.8, color = "blue")+labs(x=~"", y="", title="Ware Shoals")




garland_power_water<-left_join(Garland_Generation_Complete, gage_06279940)

garland_combined_plot <- ggplot(garland_power_water, aes(x=average_monthly_discharge, y=Power))+geom_point(alpha=0.8, color = "red")+labs(x=~"", y="", title="Garland")




chester_power_water<-left_join(Chester_Generation_Complete, gage_13046000)

chester_combined_plot <- ggplot(chester_power_water, aes(x=average_monthly_discharge, y=Power))+geom_point(alpha=0.8, color = "red")+labs(x=~"", y="", title="Chester")




tri_county_power_water<-left_join(Tri_County_Generation_Complete, gage_09146200)

tri_county_combined_plot <- ggplot(tri_county_power_water, aes(x=average_monthly_discharge, y=Power))+geom_point(alpha=0.8, color = "red")+labs(x=~"", y="", title="Tri County")




quail_power_water<-left_join(Quail_Creek_Generation_Complete, gage_09408135)

quail_combined_plot <- ggplot(quail_power_water, aes(x=average_monthly_discharge, y=Power))+geom_point(alpha=0.8, color = "red")+labs(x=~"", y="Power (MWh)", title="Quail Creek")

combined_plot<-plot_grid(hs_combined_plot, schoolfield_combined_plot, matson_combined_plot, otis_combined_plot, ws_combined_plot, garland_combined_plot, chester_combined_plot, tri_county_combined_plot,quail_combined_plot, ncol = 3, align="hv")

title_gg_3 <- ggplot() + labs(title = "Figure 3: Scatterplot of Streamflow and Power Production", subtitle = "Blue indicates eastern states while red indicates western states.")

combined_plot_grid<-plot_grid(title_gg_3, combined_plot, ncol = 1, rel_heights = c(0.08,1))+theme(plot.margin=margin(30, 10, 10, 10))

y.grob_power_2<-textGrob("Power Generation (MWh)", 
                   gp=gpar(col="black", fontsize=10), rot=90)

grid.arrange(arrangeGrob(combined_plot_grid, left = y.grob_power_2))



```

Locations of the streamflow gauges and hydroelectric plants were visualized in two distinct ways. Mapview displays the location of each hydroelectric plant, with each circle size representing the capacity of the individual plant relative to the other plants. Figure 4 uses ggplot and displays the locations of both the stream gauges and hydroelectric plants used in the analysis. 


```{r}

#Visualization of Dams

#install.packages("maps")


locations <- read.csv("./Generation_Sites.csv", stringsAsFactors = TRUE) 
locations <- locations %>%
  rename(Plant_Name = ï..Name)

final_locations<-locations[-c(9),]

states <- st_as_sf(maps::map(database = "state", plot = TRUE, fill = TRUE, col = "white"))

#Convert to a spatial dataframe
locations.sf <- final_locations %>% 
  st_as_sf(coords = c('Longitude','Latitude'),
           crs=4326)

mapview(locations.sf, cex = "MW")



gage_sites_lat_long<- read.csv("./Processed_data/gage_sites_lat_long.csv")

gage_sites_lat_long<-gage_sites_lat_long[-c(10,11), ] 

locations.gs <- gage_sites_lat_long %>% 
  rename(lat = ï..lat)%>%
  st_as_sf(coords = c('long','lat'),
           crs=4326)

ggplot()+geom_sf(data= states, fill = "white")+geom_sf(data= locations.sf)+geom_sf(data=locations.gs, col="red", alpha=0.5)+labs(title = "Figure 4: Streamflow Gauges and Hydroelectric Plant Locations")


```

Question 3: Is there a linear relationship between streamflow and power generation, and if so what is the strength of that relationship?

To investigate the linear relationship between streamflow and power generation, the lm function was utilized on each combined dataset. Each analysis resulted in a statistically significant linear regression equation and a statistically significant independet variable. The results, as indicated in Table 3, show mixed results regarding the quality of the regressions. For some hydroelectric plants, such as Tri County and Matson, greater than 40% of the variation in power generation can be explained by streamflow. However, for other plants, such as High Shoals and Quail Creek, less than 20% of the variation in power generation can be explained by streamflow. The analysis suggests that additional explanatory variables are needed to better capture fluctuations in hydroelectric generation. 


```{r}
#testing whether there is a linear relationship between streamflow and power generation 

high_shoals_regression <- lm(data = high_shoals_power_water, Power ~ average_monthly_discharge)
summary(high_shoals_regression)


schoolfield_regression <- lm(data = schoolfield_power_water, Power ~ average_monthly_discharge)
summary(schoolfield_regression)

matson_regression <- lm(data = william_f_matson_power_water, Power ~ average_monthly_discharge)
summary(matson_regression)

otis_regression <- lm(data = otis_power_water, Power ~ average_monthly_discharge)
summary(otis_regression)

ware_shoals_regression <- lm(data = ware_power_water, Power ~ average_monthly_discharge)
summary(ware_shoals_regression)

garland_regression <- lm(data =garland_power_water, Power ~ average_monthly_discharge)
summary(garland_regression)

chester_regression <- lm(data = chester_power_water, Power ~ average_monthly_discharge)
summary(chester_regression)

tri_county_regression <- lm(data = tri_county_power_water, Power ~ average_monthly_discharge)
summary(tri_county_regression)

quail_regression <- lm(data = quail_power_water, Power ~ average_monthly_discharge)
summary(quail_regression)


regression_results <- data.frame(
  Hydropower_Plants = c("High Shoals", "Schoolfield", "Matson", "Otis", "Ware Shoals", "Garland", "Chetser", "Tri County", "Quail Creek"),
  R2 = c(0.146,0.230,0.564,0.356,0.361,0.375,0.405,0.443,0.071)
)

install.packages("kableExtra")
library(kableExtra)

regression_results %>%
  kbl(caption = "Table 3: Linear Regression Results") %>%
  kable_classic(full_width = F, html_font = "Cambria")
  








```


## Results 